---
title: "Project proposal"
author: "AJJ: Armin, Jiwon, Jamari"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readxl)
library(dplyr)
```

## 1. Introduction

The data used in our group project is sourced from the OECD, specifically from the dataset “Quarterly Real GDP Growth – OECD Countries.” This dataset provides quarterly GDP growth figures for OECD member nations, covering the period from 2018 Q2 to 2025 Q2. The purpose of collecting this data is to examine how nations recovered economically following the devastating impacts of the COVID-19 pandemic, with a particular focus on comparing the varying degrees of recovery across countries. 

Our dataset from the OECD contains 1,523 data points, 31 variables, and 54 observations. We plan to exclude the “Non-OECD economies” observation since it contains no data. To fill this gap, our team is considering incorporating an additional dataset from the World Bank Group, which tracks GDP growth for both OECD and non-OECD countries. This will allow us to develop a more comprehensive and comparative view of national economic performance. We plan to later narrow in on 4 countries: USA, Denmark, South Africa, and India. We will analyze their economic performance and compare it to global post-covid performance. 

Moreover, we plan to add a few additional variables to our dataset by categorizing the levels of economic recovery into performance groups. These categories may include classifications such as stagnating, recovering, and booming. This transformation will allow us to better visualize and compare how different countries performed during the post-Covid recovery period, providing clearer insights into the relative strength and pace of their economic rebounds.

For this project, our research focuses on the varying levels of economic recovery around the world following the COVID-19 pandemic. We aim to closely examine the observations in our dataset to identify potential patterns (such as geographic or regional trends) that reveal which countries recovered quickly and which continued to face significant challenges. 

Our dataset from the OECD contains 1,523 data points, 31 variables, and 54 observations. We plan to exclude the “Non-OECD economies” observation since it contains no data. To fill this gap, our team is considering incorporating an additional dataset from the World Bank Group, which tracks GDP growth for both OECD and non-OECD countries. This will allow us to develop a more comprehensive and comparative view of national economic performance. 

For this project, our research focuses on the varying levels of economic recovery around the world following the COVID-19 pandemic. We aim to closely examine the observations in our dataset to identify potential patterns (such as geographic or regional trends) that reveal which countries recovered quickly and which continued to face significant challenges. 

Some of the questions we hope to explore include: 

* Are there clear geographic patterns indicating which countries were more effective in rebounding economically? 
* How much did the average year-on-year GDP growth rates vary across countries and continents, and were there any notable outliers?
* What factors might explain why some countries recovered more successfully than others? We plan to compare and analyze selected countries. 
* How can we categorize countries into groups such as stagnating, recovering, and booming, and what does the global distribution of these categories reveal about overall recovery trends?


## 2. Data

```{r load-data}
GDP_Growth_Filtered_Concise_Dataset_2018_2024_Annual_OECD_Non_OECD <- read_excel("/cloud/project/data/GDP Growth - Filtered Concise Dataset 2018-2024 Annual OECD + Non-OECD.xlsx")
# View(GDP_Growth_Filtered_Concise_Dataset_2018_2024_Annual_OECD_Non_OECD)

GDP_Per_Capita_2018_2024_Annual_OECD_Non_OECD <- read_excel("/cloud/project/data/GDP Per Capita - 2018 - 2024 Annual OECD + Non-OECD.xlsx")
# View(GDP_Per_Capita_2018_2024_Annual_OECD_Non_OECD)
```

```{r dimensions}
dim(GDP_Growth_Clean_Concise_Dataset_2018_2024_Annual)
dim(GDP_Per_Capita_2018_2024_Annual_OECD_Non_OECD)
```

```{r glimpse}
glimpse(GDP_Growth_Clean_Concise_Dataset_2018_2024_Annual)
glimpse(GDP_Per_Capita_2018_2024_Annual_OECD_Non_OECD)
```

## 3. Data analysis plan

The variables we will be visualizing to explore our research question are the following:

* GDP_growth: Core measure of economic recovery; allows comparison across countries and time.
* Quarter: Time-series analysis to see trends over time between 2018 Q2 to 2025 Q2.
* Country: Compare recovery across nations and regions.
* Recovery_category: Helps visualize and compare group-level performance; categorized as “Stagnating,” “Recovering,” “Booming”.
* Continent/region: Explore geographic patterns in recovery

Other additional data we may explore are datasets which will categorize the levels of economic recovery from COVID-19 into performance groups. These categories may include classifications such as stagnating, recovering, and booming. This transformation will allow us to better visualize and compare how different countries performed during the post-Covid recovery period, providing clearer insights into the relative strength and pace of their economic rebounds.

Our goal for the project is to examine the differences in economic recovery across OECD countries from the COVID-19 pandemic. We will observe using quarterly GDP growth data ranging from 2018 Q2 to 2025 Q2. Our view comes from different statistics which categorises countries that experienced the fastest and most sustained economic recoveries. 

We’ll start with exploratory data analysis to help us understand trends and the structure of our dataset. We’ll summarize tendencies, and other important information the dataset provides. Once we understand the dataset, then we’ll begin to visualize and explore the recovery patterns within the time period, while distinguishing geography, and economics.

* GDP_growth: GDP growth is the variable representing the quarterly percentage change. It is the main measure of economic recovery, and allows us to compare within countries and time periods.

* Quarter: The dataset starts in 2018 Q2 to 2025 Q2, which gives us information on GDP before, during, and after the pandemic. Looking at the times before and present day will directly provide us with the recovery time and sustainability.

* Country: We’ll use this variable to compare the economic recoveries across nations. Differences can range due to policies or resource factors that may limit a nation's growth or have it thrive.

* Recovery_category: We’ll classify countries as Stagnating, Recovering, or Booming based on their post pandemic GDP growth averages. This will allow us to visualise and for our group to make an analysis.

* Continent/Region: From this we’ll be able to identify patterns from different regions, and how different regions were affected and able to recover faster/slower.

Exploratory Data Analysis:

We’ll start by cleaning the dataset by removing unnecessary information and converting the dataset to quarterly time.
Then we’ll identify countries with the highest/lowest GDP growth past the pandemic. We can do this by using histograms and boxplots to identify outliers in the dataset.
Lastly, we’ll summarize the dataset. We’ll show who had the strongest recovery post pandemic, and how consistent the patterns were within each region:

```{r summary-statistics}

```

Planned Data Visualization:
- Line Plots: to represent the time periods 
- Box Plots: to separate data between regions 
- Bar Charts: to compare the average GDP growth by each country
- Scatter Plot: to represent the recovery time vs. GDP growth

```{r visualization}

```
