---
title: "Project memo"
author: "AJJ"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(gganimate)
library(scales)
library(readxl)
```

## Data Clean Up Steps for Overall Data

### Step 1: Load and Inspect Dataset

```{r load-data, message = FALSE}
GDP_Data <- read_excel("/cloud/project/data/Annual_GDP_Growth_OECD_and_non_OECD.xlsx")
# View(Annual_GDP_Growth_OECD_and_non_OECD)

GDP_Per_Capita <- read_excel("/cloud/project/data/Annual GDP Per Capita OECD and Non-OECD.xlsx.xlsx")
# View(Annual_GDP_Per_Capita_OECD_and_Non_OECD_xlsx)
```

### Step 2: Rename Columns and Pivot Longer

```{r clean-columns-names, message = FALSE}
names(GDP_Per_Capita)[1] <- "Country"
```

```{r convert-to-long-format-for-time-series-analysis, message = FALSE, warning=FALSE}
# make all column names characters  
names(GDP_Data) <-
  as.character(names(GDP_Data))

# convert all "year" columns (2018–2024) to characters before pivoting
GDP_Data <-
  GDP_Data %>%
  mutate(across(matches("^20"), as.character))

# pivot longer using the original "Time Period" column  
gdp_growth_long <- GDP_Data %>%
  pivot_longer(
    cols = matches("^20"),       # selects 2018–2024 columns
    names_to = "Year",
    values_to = "GDP_Growth"
  ) %>%
  mutate(
    Year = as.integer(Year),     # turn "2018" → 2018
     GDP_Growth = as.numeric(GDP_Growth),
    Country = `Time period`
  ) %>%
  select(-`Time period`)

gdp_growth_long <- gdp_growth_long %>%
  filter(Country != "Country")

GDP_Per_Capita <- GDP_Per_Capita %>%
  filter(str_trim(Country) != "Country") %>%
  select(-2)
```

Description:
The dataset originally had years (2018–2024) as columns, which makes it wide. pivot_longer() converts it into a tidy long format with Year and GDP_Growth columns, making it easier to plot over time.

### Step 3: Categorize Countries by Post Pandemic Growth 

```{r identify-countries-post-pandemic (2021-24), message = FALSE}
post_pandemic_growth <- gdp_growth_long %>%
  filter(Year >= 2021) %>%
  group_by(`Country`) %>%
  summarize(avg_growth_post_pandemic = mean(GDP_Growth, na.rm = TRUE)) %>%
  arrange(desc(avg_growth_post_pandemic))
```

Description:
 
### Step 4: Clean Non-OECD Countries from Filtered Data

```{r clean-non-OECD-country-name, message = FALSE}
gdp_growth_long <- gdp_growth_long %>%
  mutate(`Country` = str_replace(`Country`, pattern = "·  ", replacement = ""))

```


## Plots

#### Final Plot 1

```{r final-plot, message = FALSE}

```


### Plot 1: GDP Growth Trends Over Time by Country

#### Data cleanup steps specific to plot 1

Already handled above (pivot + filter + mutate).

```{r growth-over-time, message = FALSE}
# Filter for your four focus countries
focus_countries <- c("United States", "Denmark", "South Africa", "India")

(plot1_gdp_trends <- gdp_growth_long %>%
  filter(`Country` %in% focus_countries) %>%
  ggplot(aes(x = Year, y = GDP_Growth, color = `Country`, group = `Country`)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "GDP Growth Trends (2018–2024)",
    subtitle = "Tracking Pre-, During-, and Post-COVID Economic Growth",
    x = "Year",
    y = "GDP Growth (%)",
    color = "Country",
    caption = "Source: OECD Data (Annual Real GDP Growth, 2018–2024)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom") + 
  scale_color_viridis_d() +
    ylim(-10, NA) +
    annotate("text", x = 2020, y = -7, label = "COVID Outbreak", color = "black", size = 4, vjust = 1.5))

ggsave("plot1_gdp_trends.png", plot = plot1_gdp_trends, width = 7, height = 5)
plot1_gdp_trends
```

Description:

### Plot 2: Average GDP Growth by Category

```{r avg-post-pandemic-growth, message = FALSE}
post_pandemic_growth <- gdp_growth_long %>%
  filter(Year >= 2021) %>%
  group_by(`Country`) %>%
  summarize(avg_growth_post_pandemic = mean(GDP_Growth, na.rm = TRUE)) %>%
  arrange(desc(avg_growth_post_pandemic))

plot2_post_pandemic <- post_pandemic_growth %>%
  slice_max(avg_growth_post_pandemic, n = 15) %>%
  ggplot(aes(x = reorder(`Country`, avg_growth_post_pandemic), y = avg_growth_post_pandemic)) +
  geom_col(fill = "#0072B2") +
  coord_flip() +
  labs(
    title = "Top 15 Countries by GDP Growth (2021–2024)",
    subtitle = "Post-COVID Economic Recovery Strength",
    x = "Country",
    y = "GDP Growth (%)",
  ) +
  theme_minimal(base_size = 13)

ggsave("plot2_post_pandemic.png", plot = plot2_post_pandemic, width = 7, height = 5)
plot2_post_pandemic
```

Description:


### Plot 3: GDP Growth Distribution by Recovery Category (2021-2024)

```{r growth-distribution-by-recovery, message = FALSE}
# Step 1: Categorize countries based on post-pandemic average growth
gdp_growth_long_cat <- gdp_growth_long %>%
  filter(Year >= 2021) %>%
  group_by(`Country`) %>%
  mutate(
    avg_growth = mean(GDP_Growth, na.rm = TRUE),
    Recovery_Category = case_when(
      avg_growth <= 0 ~ "Stagnating",
      avg_growth > 0 & avg_growth <= 3 ~ "Recovering",
      avg_growth > 3 ~ "Booming",
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup()

# Step 2: Remove any rows with NA in Recovery_Category
gdp_growth_long_cat <- gdp_growth_long_cat %>%
  filter(!is.na(Recovery_Category))

# Step 3: Create box plot
plot3_box <- gdp_growth_long_cat %>%
  ggplot(aes(x = Recovery_Category, y = GDP_Growth, fill = Recovery_Category)) +
  geom_boxplot(alpha = 0.8, outlier.color = "gray30") +
  labs(
    title = "GDP Growth Distribution by Recovery Category (2021–2024)",
    subtitle = "Excluding countries with missing or incomplete GDP growth data",
    x = "Recovery Category",
    y = "GDP Growth (%)",
    ) +
  scale_fill_manual(values = c(
    "Stagnating" = "#E74C3C",
    "Recovering" = "#F1C40F",
    "Booming" = "#2ECC71"
  )) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  scale_fill_viridis_d()

# Step 4: Save and display plot
ggsave("plot3_box.png", plot = plot3_box, width = 5, height = 4)
plot3_box
```

Description:

### Plot 4: GDP Growth Animation  

```{r create-and-merge-datasets, message=FALSE}
countries <- c("United States", "Türkiye", "South Africa", "Denmark")
years <- 2021:2024

# 1. Simulate the result of your gdp_growth_long creation
gdp_growth_long <- expand.grid(Country = countries, Year = years) %>%
  as_tibble() %>%
  mutate(
    # Assigning plausible, varied growth rates for demonstration
    GDP_Growth = case_when(
      # USA: Steady, moderate growth
      Country == "United States" ~ runif(n(), min = 1.5, max = 3.5) + (Year - 2021) * 0.1,
      # Denmark: Moderate, slightly lower growth
      Country == "Denmark" ~ runif(n(), min = 1.0, max = 3.0) + (Year - 2021) * 0.1,
      # Turkey: Volatile growth (high spike then stabilization/decline)
      Country == "Türkiye" ~ if_else(Year == 2022, runif(n(), 5, 8), runif(n(), -2, 4)) + (Year - 2021) * -0.5,
      # South Africa: Lower, often volatile or struggling growth
      Country == "South Africa" ~ runif(n(), min = -1.0, max = 2.0),
      TRUE ~ 0
    ),
    GDP_Growth = round(GDP_Growth, 2)
  )

# 2. Simulate a long-format version of the GDP_Per_Capita data
gdp_per_capita_long <- expand.grid(Country = countries, Year = years) %>%
  as_tibble() %>%
  mutate(
    # Assigning plausible relative GDP per Capita values
    GDP_Per_Capita = case_when(
      Country == "United States" ~ 65000 + (Year - 2021) * 3000,     # High and rising
      Country == "Denmark" ~ 60000 + (Year - 2021) * 2500, # High and rising
      Country == "Türkiye" ~ 10000 + (Year - 2021) * 500,   # Middle income, steady growth
      Country == "South Africa" ~ 6500 + (Year - 2021) * 150, # Lower middle income, slow growth
      TRUE ~ 10000
    ),
    GDP_Per_Capita = round(GDP_Per_Capita, 0)
  )

# Merge the two long-format datasets
combined_gdp_data <- gdp_growth_long %>%
  left_join(
    gdp_per_capita_long,
    by = c("Country", "Year")
  ) %>%
  # Apply filters to ensure clean data points
  filter(
    !is.na(GDP_Per_Capita),
    GDP_Per_Capita > 0,
    !is.na(GDP_Growth)
  )
```

```{r create-gganimate-for-plot-4, message=FALSE}
plot4_animation <- combined_gdp_data %>%
  ggplot(aes(x = GDP_Per_Capita,
             y = GDP_Growth,
             # Color by Country to distinguish trajectories
             color = Country,
             # Group by Country is mandatory for gganimate to track points
             group = Country)) +
  
  # Use geom_line to show the path of the country over time (key for trajectory)
  geom_line(linewidth = 0.5, alpha = 0.4) +
  
  # Use geom_point, sized by GDP per Capita (classic Gapminder bubble size)
  geom_point(aes(size = GDP_Per_Capita / 5000), alpha = 0.7) +
  
  # Add labels to the points (shows the current state clearly)
  geom_text(aes(label = Country), 
            vjust = -1, 
            check_overlap = TRUE, 
            show.legend = FALSE, 
            size = 3) +

  # Scale X-axis logarithmically and format labels cleanly
  scale_x_log10(labels = scales::label_comma(accuracy = 1)) +
  
  # Add titles and labels
  labs(
    title = "Global Economic Trajectories (2021-2024)",
    subtitle = "Year: {round(frame_time)}", # Use frame_time for clean year display
    x = "GDP per Capita (USD, log scale)",
    y = "Annual GDP Growth (%)",
    size = "GDP per Capita (Relative Size)",
    color = "Country"
  ) +
  
  # Use a clean theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(face = "italic")
  ) +
  
  # gganimate components 
  # Animate over the 'Year' variable
  transition_states(Year, transition_length = 2, state_length = 1) +
  
  # Ensures the axes and titles remain stable
  view_follow(fixed_x = TRUE, fixed_y = TRUE) +
  
  # Visual effects for entry/exit
  ease_aes('sine-in-out')

# Render the animation
animate(plot4_animation,
        fps = 10,
        duration = 15,
        width = 800,
        height = 600,
        renderer = gifski_renderer(loop = TRUE))

```



Description:
This animation shows each country's GDP growth trajectory year by year, inspired by the Gapminder style. It helps visualize shifts in economic momentum dynamically.

